---
title: "Visualizing Distributions, Missing Data, Bivariate Plots"
subtitle: "PSC7475: Week 4"
author: "Prof. Weldzius"
institute: "Villanova University"
date: "Slides Updated: `r Sys.Date()`"
output:
  beamer_presentation:
    theme: "CambridgeUS"
    colortheme: "seahorse"
    fonttheme: "structurebold"
---

```{css,echo = F}
.small .remark-code { /*Change made here*/
  font-size: 85% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 50% !important;
}
```

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

---    

# Concepts and measurement
 \pause
 
  - Social science is about understanding **causal relationships** \pause 
    - Does minimum wage change levels of employment \pause
    - Does outgroup contact influence views on immigration? \pause
  
  - Relationships are between **concepts**: \pause
    - Minimum wage, unemployment, outgroup contact, views on immigration \pause
    - We took these for granted when talking about causality \pause
    
  - Important to consider how we **measure** these concepts \pause
    - Some straightforward: what is your age? \pause
    - Others more complicated: what does it mean to "be liberal"? \pause
    - **Operational definition**: mapping of concept to numbers in our data
    
---

# Example
 \pause
 
  - Concept: presidential approval  \pause
  - Conceptual definition:  \pause
    - Extent to which US adults support the actions and policies of the current US president \pause
    
  - Operational definition: \pause
    - "On a scale from 1 to 5, where 1 is least supportive and 5 is most supportive, how much would you say you support the job that Donald Trump is doing as president?"
    
---

# Measurement error
 \pause
 
  - **Measurement error**: chance variation in our measurements \pause
    - individual measurement = exact value + chance error \pause
    - chance errors tend to cancel out when we take averages \pause
    
  - No matter how careful we are, chance error can always affect a measurement. \pause
    - Panel study of 19,000 respondents: 20 reported being a citizen in 2010 and then a non-citizen in 2012 \pause
    - Data entry errors \pause
    
  - **Bias**: systematic errors for all units in the same direction. \pause
    - individual measurement = exact value + bias + chance error \pause
    - "What did you eat yesterday?" $\rightsquigarrow$ underreporting
    
---

# A biased poll?

\begin{center}
\includegraphics[width=.5\textwidth]{figs/poll_biased.png}
\end{center}  

---

# 1936 Literary Digest Poll

\begin{center}
\includegraphics[width=.45\textwidth]{figs/litdig_poll.png}
\end{center}  

 \pause
 
  - Literary Digest predicted elections using mail-in polls \pause
  - Source of addresses: automobile registrations, phone books, etc. \pause
  - In 1936, sent out 10 million ballots, over 2.3 million returned \pause
  - George Gallup used only 50,000 respondents \pause
  
  \begin{center}
  \begin{tabular}{l | r}
  & FDR's vote share \\ \hline
  Literary Digest & 43 \\
  George Gallup & 56 \\ \hline
  \end{tabular}
  \end{center}
  
---

# Poll fail

\begin{columns}
  \begin{column}{0.5\textwidth}
    \includegraphics[width=.9\textwidth]{figs/litdig_mistake.png}
  \end{column} 
  \begin{column}{0.5\textwidth}
      \begin{tabular}{l r}
      & FDR \% \\ \hline
      Literary Digest & 43 \\
      George Gallup & 56 \\ \hline
      Actual Outcome & 62 \\ 
    \end{tabular} 
  \end{column}
\end{columns}
 \pause
 
  - **Selection bias**: ballots skewed toward the wealthy (with cars, phones)  \pause
    - Only 1 in 4 households had a phone in 1936 \pause
    
  - **Nonresponse bias**: respondents differ from nonrespondents \pause
    - $\rightsquigarrow$ when selection procedure is biased, adding more units won't help!  \pause
    
---

# 1948 Election

\includegraphics[width=1\textwidth]{figs/truman.png}

---

# The Polling Disaster

\begin{center}
\begin{tabular}{l r r r r}
& Truman & Dewey & Thurmond & Wallace \\ \hline
Crossley & 45 & 50 & 2 & 3 \\
Gallup & 44 & 50 & 2 & 4 \\
Roper & 38 & 53 & 5 & 4 \\ \hline
Actual Outcome & 50 & 45 & 3 & 2
\end{tabular}
\end{center}
 \pause
 
  - **Quota sampling**: fixed quota of certain respondents for each interviewer \pause
    - If Black women make up 5% of the population, stop interviewing them once they make up 5% of your sample \pause
    
  - Sample resembles the population on these characteristics \pause
  - Potential unobserbed confounding $\rightsquigarrow$ **selection bias** \pause
  - Republicans easier to find within quotas (phones, listed addresses)
  
---

# Sample surveys
 \pause
 
  - **Probability sampling** to ensure representativeness \pause
    - Definition: every unit in the population has a known, non-zero probability of being selected into sample \pause
    
  - **Simple random sampling**: every unit has an equal selection probability. \pause
  
  - Random digit dialing: \pause
    - Take a particular area code + exchange: 310-495-XXXX. \pause
    - Randomly choose each digit in XXXX to call a particular phone \pause
    - Every phone in the US has an equal chance of being included in sample
    
---

# Sampling lingo
 \pause
 
  - **Target population**: set of people we want to learn about \pause
    - Example: people who will vote in the next election \pause
    
  - **Sampling frame**: list of people from which we will actually sample \pause
    - Frame bias: list of registered voters (frame) might include nonvoters! \pause
    
  - **Sample**: set of people contacted \pause
  
  - **Respondents**: subset of sample that acutally responds to the survey \pause
    - Unit non-response: sample $\neq$ respondents \pause
    - Not everyone picks up their phone \pause
    
  - **Completed items**: subset of questions that respondents answer \pause
    - Item non-response: refusing to disclose their vote preference
  
---

# Difficulties of sampling
 \pause
 
  - Problems of telephone survey \pause
    - Cell phones (double countring for the wealthy) \pause
    - Caller ID screening (unit non-response) \pause
    - Response rates down to 9% \pause
    
  - An alternative: internet surveys \pause
    - Opt-in panels, respondent-driven sampling $\rightsquigarrow$ **non-probability sampling** \pause
    - Cheaper, but non-representative \pause
    - Digital divide: rich vs. poor, young vs. old \pause
    - Correct for potential sampling bias via stastical methods  

---

# Effect of assasination attempts 

\pause 
\footnotesize
```{r}
library(tidyverse)
data(leaders, package = "qss")
head(leaders[,1:7])
```

---

# Contingency tables

\pause 

  - With two categorical variables, we can create **contingency tables** \pause 
    - Also known as **cross-tabs** \pause 
    - Rows are the values of one variable, columns the other \pause 

\small     
```{r}
leaders %>%
  group_by(civilwarbefore,civilwarafter) %>%
  count() %>%
  spread(civilwarafter, n)
```
\normalsize 

  - Quick summary how the two variables "go together" 

---

# Cross-tabs with proportions

```{r}
leaders %>%
  group_by(civilwarbefore,civilwarafter) %>%
  count() %>%
  ungroup() %>%
  mutate(prop = n/ sum(n)) %>%
  select(-n) %>%
  spread(civilwarafter, prop, drop = T)
```

---

# Cross-tabs with proportions (by row)

```{r}
leaders %>%
  group_by(civilwarbefore,civilwarafter) %>%
  count() %>%
  ungroup() %>%
  group_by(civilwarbefore) %>%
  mutate(prop = n/ sum(n)) %>%
  select(-n) %>%
  spread(civilwarafter, prop, drop = T)
```

---

# Scatterplot

```{r, echo=F, fig.width=4, fig.height=3}
leaders %>%
  ggplot(aes(x = politybefore, y = polityafter)) +
  geom_point(shape = 21) +
  labs(title = "Democracy before/after assasination attempts",
       x = "Democracy Level (Before)",
       y = "Democracy Level (After)") +
  theme_bw() +
  theme(plot.title = element_text(size=12))
```

---

# Scatterplot

  - Each point on the scatterplot $(x_i, y_i)$
  - Use `geom_point()` function in `ggplot`
  
```{r, echo=T, fig.show='hide'}
leaders %>%
  ggplot(aes(x = politybefore, y = polityafter)) +
  geom_point(shape = 21) +
  labs(title = "Democracy before/after assasination attempts",
       x = "Democracy Level (Before)",
       y = "Democracy Level (After)") +
  theme_bw() +
  theme(plot.title = element_text(size=12))
```

---

# Scatterplot

\footnotesize
```{r, echo=T}
leaders[1, c("politybefore","polityafter")]
```


```{r, echo=F, fig.width=3, fig.height=2}
leaders %>%
  ggplot(aes(x = politybefore, y = polityafter)) +
  geom_point(shape = 21, alpha=0) +
  geom_point(data = leaders[1, ], aes(x = politybefore, y = polityafter), color = "red", shape = 21) +
  labs(title = "Democracy before/after assasination attempts",
       x = "Democracy Level (Before)",
       y = "Democracy Level (After)") +
  theme_bw() +
  theme(plot.title = element_text(size=8),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7))
```

---

# Scatterplot

\footnotesize
```{r, echo=T}
leaders[2, c("politybefore","polityafter")]
```


```{r, echo=F, fig.width=3, fig.height=2}
leaders %>%
  ggplot(aes(x = politybefore, y = polityafter)) +
  geom_point(shape = 21, alpha=0) +
  geom_point(data = leaders[1, ], aes(x = politybefore, y = polityafter), shape = 21) +
    geom_point(data = leaders[2, ], aes(x = politybefore, y = polityafter), color = "red", shape = 21) +
  labs(title = "Democracy before/after assasination attempts",
       x = "Democracy Level (Before)",
       y = "Democracy Level (After)") +
  theme_bw() +
  theme(plot.title = element_text(size=8),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7))
```

---

# Scatterplot

\footnotesize
```{r, echo=T}
leaders[3, c("politybefore","polityafter")]
```


```{r, echo=F, fig.width=3, fig.height=2}
leaders %>%
  ggplot(aes(x = politybefore, y = polityafter)) +
  geom_point(shape = 21, alpha=0) +
  geom_point(data = leaders[1, ], aes(x = politybefore, y = polityafter), shape = 21) +
  geom_point(data = leaders[2, ], aes(x = politybefore, y = polityafter), shape = 21) +
  geom_point(data = leaders[3, ], aes(x = politybefore, y = polityafter), color = "red", shape = 21) +
  labs(title = "Democracy before/after assasination attempts",
       x = "Democracy Level (Before)",
       y = "Democracy Level (After)") +
  theme_bw() +
  theme(plot.title = element_text(size=8),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7))
```


---

# Scatterplot

\footnotesize
```{r, echo=T}
leaders[3, c("politybefore","polityafter")]
```


```{r, echo=F, fig.width=3, fig.height=2}
leaders %>%
  ggplot(aes(x = politybefore, y = polityafter)) +
  geom_point(shape = 21) +
  labs(title = "Democracy before/after assasination attempts",
       x = "Democracy Level (Before)",
       y = "Democracy Level (After)") +
  theme_bw() +
  theme(plot.title = element_text(size=8),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7))
```

---

# How big is big?

\pause 
  - Would be nice to have a standard summary of how similar variables are \pause
    - Problem: variables on different scales! \pause
    - Needs a way to put any variable on common units \pause
    
  - **z-score** to the rescue! \pause
  
  $$
  \text{z-score of } x_i = \frac{x_i - \text{mean of } x}{\text{standard deviation of }x}
  $$
 \pause
 
  - Crucial property: z-scores don't depend on units \pause
  
  $$
  \text{z-score of } (ax_i + b) = \text{z-score of } x_i
  $$
  
---

# Correlation
 \pause
 
  - How do variables move together on average? \pause
  - When $x_i$ is big, what is $y_i$ likely to be? \pause
    - Positive correlation: when $x_i$ is big, $y_i$ is also big \pause
    - Negative correlation: when $x_i$ is big, $y_i$ is small \pause
    - High magnitude of correlation: data cluster tightly around a line \pause
    
  - The technical definition of the **correlation coefficient**: \pause
  
  $$
  \frac{1}{n-1} \sum_{i=1}^{n}{\left[ (\text{z-score for }x_i)\times(\text{z-score for }y_i) \right]}
  $$
  
---

# Correlation intuition:
 
\begin{center}
\includegraphics[width=.75\textwidth]{figs/corr1.png}
\end{center}

---

# Correlation intuition:

 
\begin{center}
\includegraphics[width=.75\textwidth]{figs/corr2.png}
\end{center}
\pause

  - Large values of $X$ tend to occur with large values of $Y$ \pause
    - (z-score for $x_i$) $\times$ (z-score for $y_1$) = (pos. num.) $\times$ (pos. num.) $= +$ \pause
  - Small values of $X$ tend to occur with small values of $Y$   \pause
    - (z-score for $x_i$) $\times$ (z-score for $y_1$) = (neg. num.) $\times$ (neg. num.) $= +$ \pause
  - If these dominate $\rightsquigarrow$ positive correlation
  
---

# Correlation intuition:

 
\begin{center}
\includegraphics[width=.75\textwidth]{figs/corr3.png}
\end{center}
\pause

  - Large values of $X$ tend to occur with small values of $Y$ \pause
    - (z-score for $x_i$) $\times$ (z-score for $y_1$) = (pos. num.) $\times$ (neg. num.) $= -$ \pause
  - Small values of $X$ tend to occur with large values of $Y$   \pause
    - (z-score for $x_i$) $\times$ (z-score for $y_1$) = (neg. num.) $\times$ (pos. num.) $= -$ \pause
  - If these dominate $\rightsquigarrow$ negative correlation  
    
---

# Properties of correlation coefficient

 \pause
 
  - Correlation measures **linear** association. \pause
  - Interpretation: \pause
    - Correlation is between -1 and 1 \pause
    - Correlation of 0 means no linear association \pause
    - Positive correlations $\rightsquigarrow$ positive associations \pause
    - Negative correlations $\rightsquigarrow$ negative associations \pause
    - Closer to -1 or 1 means stronger association \pause
  - Order doesn't matter: `cor`(x,y) = `cor`(y,x) \pause
  - Not affected by changes of scale: \pause
    - `cor`(x,y) = `cor`(ax+b, cy+d)
    - Celsius vs. Fahrenheit; dollars vs. pesos; cm vs. in.
    
---

# Correlation in R

  - Use the `cor()` function \pause 
  
```{r}
leaders %>%
  select(politybefore, polityafter) %>%
  cor()
```
  
  - Very highly correlated!

---

# Assassination attempts

  - See the possible attempt results
  
```{r}
unique(leaders$result)
```

--- 

# Creating an attempt fatal variable

  - use `ifelse` to create a `fatal` variable

```{r}
## create new vector of unique results of "result" 
lev <- unique(leaders$result)
leaders <- leaders %>%
  mutate(fatal = ifelse(result %in% lev[c(2,7,8,10)], 1,0))
leaders %>%
  summarize(mean(fatal))
```

---

# Remember boxplots?

```{r, echo=F, fig.width=3.5, fig.height=2.5}
leaders %>%
  ggplot(aes(y = politybefore)) + 
  geom_boxplot() +
  scale_y_continuous(breaks = seq(-10, 10,by = 5)) + 
  xlim(-.75,.75) + 
  labs(title = "Pre-Attempt Democracy Level", y = "Pre-Attempt Polity Score") +
  theme_bw() + 
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

---

# Comparing distribution with the boxpot

  - What if we want to know how the distribution varies by success?

```{r, echo=F, fig.width=3.5, fig.height=2.5}
leaders %>%
  ggplot(aes(y = politybefore, x = factor(fatal,labels = c("Survived", "Died")))) + 
  geom_boxplot() +
  scale_y_continuous(breaks = seq(-10, 10,by = 5)) + 
  labs(title = "Pre-attempt Democracy Level by Attempt Outcome", y = "Pre-Attempt Polity Score", x = "Assassination Attempt Outcome") +
  theme_bw() +
  theme(plot.title = element_text(size=9),
        axis.title.x = element_text(size = 9),
        axis.title.y = element_text(size = 9),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

---

# Boxplot comparisons in R

```{r, echo=T, fig.show='hide'}
leaders %>%
  ggplot(aes(y = politybefore, 
             x = factor(fatal,labels = c("Survived", "Died")))) + 
  geom_boxplot() +
  scale_y_continuous(breaks = seq(-10, 10,by = 5)) + 
  labs(title = "Pre-attempt Democracy Level by Attempt Outcome", 
       y = "Pre-Attempt Polity Score", 
       x = "Assassination Attempt Outcome") +
  theme_bw() +
  theme(plot.title = element_text(size=9),
        axis.title.x = element_text(size = 9),
        axis.title.y = element_text(size = 9),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

---

# Quantile-Quantile Plot

  - How do we compare distributions of two variables that are not in the same dataset?
    - Could use boxplots, but it's only a crude summary of the distributions.
  - **Quantile-quantile plot (Q-Q plot)**: scatterplot of **quantiles**
    - (min of $X$, min of $Y$)
    - (median of $X$, median of $Y$)
    - (25th percentile of $X$, 25th percentile of $Y$)
  - Intuitions:
    - If distributions are the same $\rightsquigarrow$ all points on a 45-degree line
    - Points above 45$^{\circ}$ line $\rightsquigarrow$ y-axis variable has larger value of the quantile
    - Point below 45$^{\circ}$ line $\rightsquigarrow$ x-axis variable has larger value of the quantile
    - Steeper slope than 45$^{\circ}$ line $\rightsquigarrow$ y-axis variable has more spread
    - Flatter slope than 45$^{\circ}$ line $\rightsquigarrow$ x-axis variable has more spread
    
---

# QQ-plot example

```{r, echo=F, fig.width=4, fig.height=3}
leaders <- leaders %>%
  mutate(polity_change = polityafter - politybefore,
         polity_change_f = ifelse(fatal==1,polity_change,NA),
         polity_change_nf = ifelse(fatal==0,polity_change,NA))

quantile_probs <- seq(from = 0, to = 1, by = 0.01)
quantile_names <- as.character(quantile_probs)

quantiles <- leaders %>%
  group_by(fatal) %>%
  summarize(politychng_quantile = quantile(polity_change, probs = quantile_probs),
            quantile = quantile_names) %>%
  pivot_wider(names_from = fatal,
              values_from = politychng_quantile) 
  

quantiles %>%
  ggplot(aes(x = `0`, 
             y = `1`)) + 
  geom_point(shape = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  scale_y_continuous(breaks = seq(-20, 15,by = 5)) +
  scale_x_continuous(breaks = seq(-20, 15,by = 5)) +
  labs(title = "Change in Polity: Leader Survived vs. Leader Died", 
       y = "Change in Polity, Leader Died",
       x = "Change in Polity, Leader Died") +
  theme_bw() +
  theme(plot.title = element_text(size=9),
        axis.title.x = element_text(size = 9),
        axis.title.y = element_text(size = 9))
```

---

# QQ-plot example (setup)

\footnotesize 
```{r, echo=T, results='hide'}
## calculate change in polity
leaders <- leaders %>%
  mutate(polity_change = polityafter - politybefore)

## set quantile vectors
quantile_probs <- seq(from = 0, to = 1, by = 0.01)
quantile_names <- as.character(quantile_probs)

## generate dataframe for plot
quantiles <- leaders %>%
  group_by(fatal) %>%
  summarize(politychng_quantile = quantile(polity_change, probs = quantile_probs),
            quantile = quantile_names) %>%
  pivot_wider(names_from = fatal,
              values_from = politychng_quantile) 
```


---

# QQ-plot example (plot)

\small 
```{r, echo=T, fig.show='hide'}
quantiles %>%
  ggplot(aes(x = `0`, y = `1`)) + 
  geom_point(shape = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  scale_y_continuous(breaks = seq(-20, 15,by = 5)) +
  scale_x_continuous(breaks = seq(-20, 15,by = 5)) +
  labs(title = "Change in Polity: Leader Survived vs. Leader Died", 
       y = "Change in Polity, Leader Died",
       x = "Change in Polity, Leader Died") +
  theme_bw() +
  theme(plot.title = element_text(size=9),
        axis.title.x = element_text(size = 9),
        axis.title.y = element_text(size = 9))
```



  
    