---
title: "The End of Currency Manipulation? "
subtitle: "Global Production Networks and Exchange Rate Outcomes"
author: "Ryan Weldzius"
institute: "Villanova University"
date: "Slides Updated: `r Sys.Date()`"
output:
  beamer_presentation:
    theme: "CambridgeUS"
    colortheme: "seahorse"
    fonttheme: "structurebold"
header-includes:
  - \usepackage{tikz}
  - \usepackage{graphicx} 
  - \usepackage{amssymb}
  - \usepackage{fontenc}
  - \usepackage{xcolor}    
  - \usepackage{array}
# classoption: "handout"  
---

```{css,echo = F}
.small .remark-code { /*Change made here*/
  font-size: 85% !important;
}
.tiny .remark-code { /*Change made here*/
  font-size: 50% !important;
}
```

```{r,include=F}
set.seed(123)
options(width=60)
knitr::opts_chunk$set(fig.align='center',fig.width=9,fig.height=5,message=F,warning=F)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
library(tidyverse)
library(tinytex)
library(gridExtra)
```

# Why study currency manipulation?

- Historical evidence that this strategy works \pause 
  - Taiwan \pause
  - Germany \pause
  - Japan (pre-1985, Plaza Accord) \pause
  - China (1995-2005) \pause
  - South Korea, Sweden, Czechia, Israel, Vietnam, Malaysia, Indonesia \pause
- 2018: No manipulators remained; \pause why?

---

# What caused this "end" of manipulation?

- Market competitiveness; no longer benefits them \pause
- Production plateau - can't keep up with demand \pause
- Trump ends it \pause
- IMF policy \pause
- My argument: internationalization of production

---

# Currency manipulation

```{r, echo=FALSE, fig.height=3, fig.width=4}
# echo = FALSE stops Rmarkdown from showing anything in this chunk when you knit 
# fig.height and fig.width might need to be adjusted

df <- read_csv("~/Downloads/currencymanip2020.csv")

df <- df %>%
  mutate(forex_gdp = lag(exp(l_forex_gdp_f1))) %>%
  filter(country!="Libya") %>%
  group_by(year) %>%
  mutate(forex_gdp_mean = mean(forex_gdp, na.rm=T)) %>%
  ungroup()

df %>%
  ggplot(aes(x=year, y=forex_gdp, group=country)) +
  geom_line(alpha=0) +
  geom_line(aes(x=year, y=forex_gdp_mean), size=1, color="red", alpha=0)+
  theme_bw() +
  labs(x="", y="Forex/GDP (%)") +
  scale_x_continuous(limits=c(1990,2018), breaks=seq(1990,2018,5))
```


---

# Currency manipulation

```{r, echo=FALSE, fig.height=3, fig.width=4}
# echo = FALSE stops Rmarkdown from showing anything in this chunk when you knit 
# fig.height and fig.width might need to be adjusted
df %>%
  ggplot(aes(x=year, y=forex_gdp, group=country)) +
  geom_line(alpha=0.15) +
  geom_line(aes(x=year, y=forex_gdp_mean), size=1, color="red", alpha=0)+
  theme_bw() +
  labs(x="", y="Forex/GDP (%)") +
  scale_x_continuous(limits=c(1990,2018), breaks=seq(1990,2018,5))
```

---

# Currency manipulation

```{r, echo=FALSE, fig.height=3, fig.width=4}
# echo = FALSE stops Rmarkdown from showing anything in this chunk when you knit 
# fig.height and fig.width might need to be adjusted
df %>%
  ggplot(aes(x=year, y=forex_gdp, group=country)) +
  geom_line(alpha=0.15) +
  geom_line(aes(x=year, y=forex_gdp_mean), size=1, color="red")+
  theme_bw() +
  labs(x="", y="Forex/GDP (%)") +
  scale_x_continuous(limits=c(1990,2018), breaks=seq(1990,2018,5))
```

---

# Currency manipulation and Global Value Chains

```{r, echo=FALSE, fig.height=3, fig.width=4}
# echo = FALSE stops Rmarkdown from showing anything in this chunk when you knit 
# fig.height and fig.width might need to be adjusted
df %>%
  mutate(gvc_exp = exp(l_gvc_exp)) %>%
  filter(gvc_exp<5) %>%
  ggplot(aes(x=gvc_exp, y=forex_gdp)) +
  geom_point(alpha=0) +
  theme_bw() +
  labs(x="GVC/Exports (%)", y="Forex/GDP (%)")
```

---

# Currency manipulation and Global Value Chains

```{r, echo=FALSE, fig.height=3, fig.width=4}
# echo = FALSE stops Rmarkdown from showing anything in this chunk when you knit 
# fig.height and fig.width might need to be adjusted
df %>%
  mutate(gvc_exp = exp(l_gvc_exp)) %>%
  filter(gvc_exp<5) %>%
  ggplot(aes(x=gvc_exp, y=forex_gdp)) +
  geom_point(alpha=0.15) +
  theme_bw() +
  labs(x="GVC/Exports (%)", y="Forex/GDP (%)")
```

---

# Currency manipulation and Global Value Chains

```{r, echo=FALSE, fig.height=3, fig.width=4}
# echo = FALSE stops Rmarkdown from showing anything in this chunk when you knit 
# fig.height and fig.width might need to be adjusted
df %>%
  mutate(gvc_exp = exp(l_gvc_exp)) %>%
  filter(gvc_exp<5) %>%
  ggplot(aes(x=gvc_exp, y=forex_gdp)) +
  geom_point(alpha=0.15) +
  theme_bw() +
  labs(x="GVC/Exports (%)", y="Forex/GDP (%)") +
  geom_smooth(method="lm")
```

---

# Hypotheses

- H1: more dependent on trade, the more likely to manipulate \pause
- H2: more dependent on trade, **conditional** on global value chain dependence, the **less** likely to manipulate \pause 

\begin{align*}
\text{forex/gdp}_{it} = & \alpha + \beta_1 (\text{Exports/GDP})_{it} + \beta_2 (\text{GVC/Exports})_{it} \\ 
& + \beta_3 (\text{Exports/GDP})_{it} \times (\text{GVC/Exports})_{it} + \epsilon_{it}
\end{align*}

- $i$ indicates the country
- \emph{t} indicates the year

---

# What is the affect of GVC trade on currency manipulation?

```{r, echo=FALSE, message=FALSE}
# Here you can include your statistical model. You'll want to output the results in a readable way. 
# Use this link to help with formatting these tables (see "regression table"); note that you'll have to use some new libraries (and note what they include at the beginning of the R chunk, e.g., "message = FALSE"): https://zief0002.github.io/book-8252/pretty-printing-tables-in-markdown.html

df$year_factor = as.factor(df$year)
df$country_factor = as.factor(df$country)
library(lfe)

m1 <- felm(l_forex_gdp_f1 ~ l_exp_goods_gdp   | year_factor + country_factor | 0 | country_factor, data=df)
m2 <- felm(l_forex_gdp_f1 ~ l_exp_goods_gdp*l_gvc_exp   | year_factor + country_factor | 0 | country_factor, data=df)

```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis'} 
library(stargazer)
table <- stargazer(m1, m2, 
          type = "latex",
          header = FALSE,
          no.space = TRUE,
          float = FALSE,
          keep.stat = c("n","adj.rsq"), # only keep what I need
          column.labels = c("Model 1", "Model 2"),
          colnames = FALSE,
          model.numbers = FALSE,
          dep.var.caption = " ",
          dep.var.labels = "",
          covariate.labels = c("Exports/GDP", "GVC/Exports","(Exp/GDP)x(GVC/Exp)"),
          notes.align = "l"
)

#table <- table[-c(3:7)] # removes the top lines of the Stargazer table so we can see more rows
cat(paste(table, collapse = "\n"))
```


---

# Slide 5 title (concluding thoughts/next steps)

- Layout your next steps (the `\pause` after each line allows me to reveal each line with a click; no need to put this on your last line of text): \pause
  1. I need to look at the impact of a particular covariate/control, or add more controls \pause
  2. Check residuals \pause
  3. etc.


---

# Appendix 1

```{r, echo=FALSE, fig.height=3, fig.width=4}
# echo = FALSE stops Rmarkdown from showing anything in this chunk when you knit 
# fig.height and fig.width might need to be adjusted
df %>%
  ggplot(aes(x=forex_gdp)) +
  geom_density() 
```

---

# Appendix 2

```{r, echo=FALSE, fig.height=3, fig.width=4}
# echo = FALSE stops Rmarkdown from showing anything in this chunk when you knit 
# fig.height and fig.width might need to be adjusted
df %>%
  ggplot(aes(x=l_forex_gdp_f1)) +
  geom_density() 
```